---
title: 'Factor Analysis - Advanced Measurement Models'
teaching: 10
exercises: 15
---

:::::::::::::::::::::::::::::::::::::: questions 

- What is a bifactor model?
- What is a hierachical model?
- How can I specify different types of measurement models?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Understand the difference between hierarchical, non-hierarchical and bifactor models
- Learn how to specify a hierarchical model
- Learn how to specify a bifactor model
- Learn how to evaluate hierachical and bifactor models
::::::::::::::::::::::::::::::::::::::::::::::::

## Measurement Models

Something about MMS in general

Nested models

Hieraschical
- exmaple uses and specifications

Bifactor
- example uses and specifcations

```{r}
model_cfa_bifac <- c(
  "
  # Model Syntax
  
  # =~ is like in linear regression
  # left hand is latent factor, right hand manifest variables contributing
  
  # Define which items load on which factor
  depression =~ Q3A + Q5A + Q10A + Q13A + Q16A + Q17A + Q21A + Q24A + Q26A + Q31A + Q34A + Q37A + Q38A + Q42A
  
  anxiety =~ Q2A + Q4A + Q7A + Q9A + Q15A + Q19A + Q20A + Q23A + Q25A + Q28A + Q30A + Q36A + Q40A + Q41A
  
  stress =~ Q1A + Q6A + Q8A + Q11A + Q12A + Q14A + Q18A + Q22A + Q27A + Q29A + Q32A + Q33A + Q35A + Q39A
  
  g_dass =~ Q3A + Q5A + Q10A + Q13A + Q16A + Q17A + Q21A + Q24A + Q26A + Q31A + Q34A + Q37A + Q38A + Q42A +Q2A + Q4A + Q7A + Q9A + Q15A + Q19A + Q20A + Q23A + Q25A + Q28A + Q30A + Q36A + Q40A + Q41A + Q1A + Q6A + Q8A + Q11A + Q12A + Q14A + Q18A + Q22A + Q27A + Q29A + Q32A + Q33A + Q35A + Q39A
  
  # Define correlations between factors using ~~
  depression ~~ anxiety
  depression ~~ stress
  anxiety ~~ stress
  
  g_dass ~~ 0*depression
  g_dass ~~ 0*stress
  g_dass ~~ 0*anxiety
  "
)
```

```{r}
fit_cfa_bifac <- cfa(model_cfa_bifac, data = fa_data)

summary(fit_cfa_bifac, fit.measures = TRUE, standardized = TRUE)
```



Declaring a winner for our DASS data

Important note, do not confuse confirmatory with exploratory research!


::: challenge 

## Challenge 1

Read in data and answer the question: do we have a hypothesis as to how the structure should be?

:::

::: challenge 

## Challenge 2

Run a CFA with three factors
:::

::: challenge 

## Challenge 3

Interpret some results
:::

::: challenge 

## Challenge 4

Run a cfa with one factor, interpret some results

:::

::: challenge 

## Challenge 5

Compare the models from challenge 3 and challenge 4, which one fits better?

:::

::::::::::::::::::::::::::::::::::::: keypoints 

- Something

::::::::::::::::::::::::::::::::::::::::::::::::

